{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory

### 2023-10-29 paper/si.tex
remove proposition 1 and tildes

\noindent
This document provides further details on our methods and derivations, and it contains additional simulation results. 
Section~\ref{section:model} summarizes the model. 
In particular, we provide further details on our implementation of the evolutionary dynamics, and our use of the rare-mutations limit. 
In Section~\ref{section:analyticalresults} we derive analytical results for the various settings we consider.
These settings differ in what kind of payoff information individuals take into account when updating their strategies. 
In the perfect memory setting, individuals take into account all their interactions against all co-players.
In the limited memory setting, they only consider the very last round of their very last interaction.  
In addition, we describe several model extensions in which the amount of information taken into account is in between these two extremes. 
Finally, Section~\ref{section:furthersimulations} presents further simulation results. 
In particular, we confirm that our main results continue to hold ({\it i}) when mutations are no longer rare, and  ({\it ii}) when players use memory-1 strategies instead of reactive strategies.

@@

\noindent
{\bf Summary of the model.} As described in the main text, we study cooperative behavior in a population of size $N$, with $N$ being even.
The dynamics unfolds on two time-scales. 
The short time scale describes the game dynamics. 
Here the $N$ individuals are randomly matched to form $N/2$ pairs to interact in a repeated prisoner's dilemma.  
Each round, individuals can choose whether to cooperate ($C$) or defect ($D$). 
In the most general setting, the resulting one-shot payoffs can be summarized by the payoff matrix 
\begin{align}
	\bordermatrix{%
		& C & D \cr
		C &\ R &\ S\  \cr
		D &\ T &\ P\ \cr
	} . \label{eq:prisoner}
\end{align}
Here, $R$ is interpreted as the reward for mutual cooperation, $S$ is the sucker's payoff, $T$ is the temptation, and $P$ is the punishment payoff~\citep{axelrod1981evolution}. 
Throughout this work, we parametrize these payoffs as $R\!=\!b\!-\!c$, $S\!=\!-c$, $T\!=\!b$, and $P\!=\!0$, where $b$ and $c$ are the benefit and cost of cooperation, respectively, with $b\!>\!c\!>0$. 
After each round, players learn their co-player's previous action. 
Then the game continues for another round with probability~$\delta$.
Players make their decisions whether to cooperate in any given round based on their reactive strategies $\mathbf{s}\!=\!(y,p,q)$. 
The entry $y$ determines a player's first-round cooperation probability. 
The other entries $p$ and $q$ determine the player's cooperation probability in all subsequent rounds. 
The probability is $p$ if the co-player cooperated in the previous round, and it is $q$ if the co-player defected. 
On this short time scale, the players' strategies are fixed, and players are consecutively matched to play many repeated games with randomly changing interaction partners.

@@

The long time scale describes the evolutionary dynamics. 
Here, players are allowed to update their strategies based on the payoffs they yield. 
We model these strategy updates with a pairwise comparison process~\citep{traulsen2007pairwise}.
This process assumes that at regular time intervals, one player is randomly selected from the population.
We refer to this player as the `learner' (\learner). 
The learner is then given an opportunity to update its strategy. 
There are two possibilities for how this update may occur. 
With probability $\mu$, the player's strategy mutates randomly. 
In that case, the player's new strategy is drawn uniformly from the space of all reactive strategies~$[0,1]^3$.  
With probability $1\!-\!\mu$, the player engages in a pairwise comparison. 
In that case, the player randomly picks another individual from the population (referred to as the `role model', \rolemodel). 
The learner adopts the role model's strategy with a probability \(\rho\) given by
\begin{equation} \label{Eq:rho}
    \rho(\pi_{L}, \pi_{R}) = \frac{1}{1\!+\! e^{\!-\!\beta (\pi_\rolemodel - \pi_\learner)}}.
\end{equation}
The parameter \(\beta\) is the selection strength.
It determines how important payoff differences are for the learner's decision to imitate the role model. 
The variables $\pi_\rolemodel$ and $\pi_\learner$ refer to the relevant payoffs of the role model and the learner, respectively. 
The exact value of these payoffs depend on the players' memory. 
We say players have {\it perfect memory} when $\pi_\rolemodel$ and $\pi_\learner$ are given by the players' expected payoffs (across all rounds and across all possible co-players). 
We say players have {\it limited memory} when $\pi_\rolemodel$ and $\pi_\learner$ are given by the players' realized payoff in the very last round of the game with their very last interaction partner. 
In addition, we consider several model extensions in which individuals have memory capacities in between these two extremes. 
We provide a detailed description of these different settings and the resulting payoffs in Section~\ref{section:analyticalresults}.\\

%% MODEL: Intro to the rare-mutations limit %%

\noindent
{\bf Evolutionary simulations for the rare-mutations limit.} 
To simulate the evolutionary dynamics of the pairwise comparison process, it is sometimes useful to assume that mutations are rare, $\mu\!\rightarrow\!0$. 
In that case, whenever a mutant strategy appears, it either fixes in the population or goes extinct before the next mutant appears. 
As a result, at any given time there are at most two different strategies present in the population~\citep{fudenberg:JET:2006,wu:JMB:2012,mcavoy:jet:2015}. 
This assumptions makes computations more efficient, and it makes some of the results easier to interpret.
In the following, we describe our implementation of the process in the rare-mutations limit in more detail.

%% MODEL: Implementation of the rare-mutations limit %%

Initially, the process starts with a population where all members use the same strategy (referred to as the resident strategy~\resident). 
Then one individual adopts a mutant strategy selected uniformly at random from the set of feasible strategies.
The fixation probability \(\varphi_{\mutant}\) of the mutant strategy can be calculated explicitly~\citep{nowak:Nature:2004},
\begin{equation}\label{eq:fixation_probability}
    \varphi_{\mutant} = \frac{1}{1+\sum\limits_{i=1}^{N-1}\prod\limits_k^i \frac{\lambda^-_k}{\lambda^+_k}}.
\end{equation}
Here, the index $k$ corresponds to the current number of players with the mutant strategy (mutants). 
The variables \(\lambda^-_k, \lambda^+_k\) are the probabilities that the number of mutants decreases or increases within a single updating step. 
These probabilities depend on the probability that a mutant and a resident are chosen as the learner and the resident, respectively. 
In addition, they depend on the respective switching probability~$\rho$, as described by Eq.~\eqref{Eq:rho}. 
We specify the exact values of  \(\lambda^-_k, \lambda^+_k\) for each memory-setting in the next section.

Depending on the fixation probability \(\varphi_{\mutant}\), the mutant strategy either fixes (becomes the new resident) or goes extinct. 
Afterwards, another random mutant strategy is introduced into the population. 
We iterate this elementary population updating process for a large number of mutant strategies. 
At each step, we record the current resident strategy, and the resulting average cooperation rate.

We consider this limit of rare-mutations throughout the main text. 
The respective process is summarised by Algorithm~\ref{algorithm:pairwise_comparison}.
In Section~\ref{section:furthersimulations} we present additional simulation results to show that our qualitative results continue to hold when the mutation rate is strictly bounded away from zero.

@@

Using the shortcut notation $r_i:=p_i\!-\!q_i$, we obtain

@@

~\\
{\bf Stochastic stability of cooperation.}
As an application of this formalism, we can compute when cooperation is stochastically stable in the perfect information setting. 
To this end, suppose there is only a single mutant, $k\!=\!1$. The residents adopt Generous Tit-for-Tat, $\gtft = (1, 1, q)$ and the mutant adopts $\alld\! = \! (0,0,0)$. When two GTFT players interact, the resulting average distribution according to Eq.~\eqref{Eq:vAverage} simplifies to
\begin{align*}
\mathbf{v}(\gtft,\gtft) = (1, 0, 0, 0).
\end{align*}
On the other hand, if \alld{} interacts with \gtft, the respective probabilities become,
\begin{align*}
  \mathbf{v}(\alld,\gtft) = \big(0,~0,~1\!-\!\delta\!+\!\delta q,~\delta(1\! -\! q)\big).
\end{align*}
Based on Eq.~\eqref{Eq:ExpPay}, we can compute the strategies' expected payoffs as
\begin{align*}
\pi_\gtft =  \frac{N\!-\!2}{N-1} (b - c)  \!-\! \frac{1}{N-1}\big(1\!-\!\delta\!+\!\delta q\big)c
\quad \text{and} \quad
\pi_\alld  = \big(1\!-\!\delta\!+\!\delta q\big)b.
\end{align*}
As a consequence, we can calculate the corresponding ratio of transition probabilities according to Eq.~\eqref{eq:LambdaRatio},
\begin{align*}
\frac{\lambda_1^{-}}{\lambda_1^{+}} = e^{-\beta \left( (1-\delta+\delta q)(b+\frac{c}{N-1})- \frac{N-2}{N-1}(b-c) \right)}
\end{align*}
By definition, cooperation is stochastically stable if this ratio exceeds one, which is equivalent to
\begin{equation*}
q<1-\frac{1}{\delta}\cdot \frac{b+(N\!-\!1)c}{(N\!-\!1)b+c}.
\end{equation*}
For such a strategy to be feasible we require $q\!>\!0$, which implies $\delta \!>\! \big(b+(N\!-\!1)c\big)/\big((N\!-\!1)b\!+\!c\big)$.
In particular, in the limit of large populations \(N \!\rightarrow\! \infty\), we obtain that cooperation is stochastically stable if  \(q \!<\! 1\! -\! c/(\delta b)\). 
The minimum continuation probability for such a strategy to exist is $\delta \!>\! c/b$. 
In this way, we recover the classical conditions for cooperation to be feasible under direct reciprocity~\citep{molander:jcr:1985,Nowak1992tit,Schmid:NHB:2021}.

@@

{\bf Computing the distribution of last-round payoffs.}
The case of perfect payoff memory is straightforward to handle; here, every player gets the expected payoff with certainty. 
In comparison, computing transition probabilities for the case of limited payoff memory is more elaborate. 
Here, we need to consider the different possible outcomes that both the learner and role model may have experienced in their very last interaction. 
A further complication arises when the learner's last interaction partner happens to be the role model. 
In that case, the learner's and the role model's last payoff will be correlated (e.g., if the learner got the sucker's payoff of $S$, the role model's payoff is $T$ with certainty). 
To treat the case of limited memory analytically, let $\mathcal{U} \!=\! \{R,S,T,P\}$ be the set of possible one-shot payoffs. 
We first  compute the probability that a player receives a payoff of $u\!\in\!\mathcal{U}$ in the final round of the repeated game when the strategies of the two players are given.

\begin{Prop}\label{proposition:last_round} 
Consider a repeated game with continuation probability $\delta$, between two players with reactive strategies $\strategy_1\!=\!(y_1, p_1, q_1$)  and $\strategy_2\!=\!(y_2,p_2,q_2)$. 
Let $\tilde{v}_u\!=\!\tilde{v}_u(\strategy_1,\strategy_2)$ denote the probability that the first player receives the payoff $u\!\in\!\mathcal{U}$ in the final round of the game. 
Then this probability  $\tilde{v}_u$ is identical to the player's average probability $v_u$ to receive payoff $u$ across all rounds of the game, as specified by Eq.~\eqref{Eq:vAverage}.
\end{Prop}

\begin{proof}
Let $\mathbf{\tilde{v}}:=(\tilde{v}_R, \tilde{v}_S, \tilde{v}_T, \tilde{v}_P)$. 
To compute $\mathbf{\tilde{v}}$, we first compute the probability to receive each outcome $u\!\in\!\mathcal{U}$ given that the last round happens to be round $\tau\!\in\!\{0,1,\ldots\}$. 
In that case, it follows from Eqs.~\eqref{Eq:v0} and~\eqref{Eq:vt} that the respective probability is
\begin{equation} \label{Eq:vTildeSimple}
\mathbf{\tilde v}(\tau) = \mathbf{v_0}\cdot M^\tau.
\end{equation}
Round $\tau$ is the last round with probability $\delta^\tau(1\!-\!\delta)$. 
By the law of total probability, we can obtain $\mathbf{\tilde{v}}$ by summing up over all possible last rounds,
\begin{equation}
\tilde{v} = \sum_{\tau=0}^\infty \delta^\tau(1\!-\!\delta)  \cdot \mathbf{v_0}M^\tau 
=  (1\!-\!\delta)\mathbf{v_0} \sum_{t=0}^\infty \delta^t  M^t. 
\end{equation}
This expression for $\mathbf{\tilde{v}}$ coincides with the expression for $\mathbf{v}$ in Eq.~\eqref{Eq:vDefinition}.
\end{proof}

\noindent
Note that while we focus on the prisoner's dilemma in this work, the Proposition applies to any repeated matrix game.\\

@@

\noindent
{\bf Computing the ratio $\lambda^-_k/\lambda^+_k$.}
After these preparations, let us again consider the corresponding population setup, with $N\!-\!k$ residents with strategy  \(\strategy_\resident = (y_\resident, p_\resident, q_\resident)\) and $k$ mutants with strategy  \(\strategy_\mutant = (y_\mutant, p_\resident, q_\mutant)\).
At each step of the evolutionary process we choose a learner and a role model. 
The learner compares the performance of its strategy by comparing its own last one-shot payoff with the last one-shot payoff of the role model. 
In the following, we assume either the learner or the role model is a resident and that the other player is a mutant (otherwise it is certain that the number of mutants does not change). 
There are two major cases to consider.

@@

\begin{enumerate}
\item The learner and the role model have their last respective interaction with each other. This happens with probability $1/(N\!-\!1)$. In that case, there are four possible cases for their joint final payoffs, $(u_\resident,u_\mutant)\!\in\!\mathcal{U}_F\!:=\!\big\{(R,R),\,(S,T),\,(T,S)\,(P,P)\big\}$. These outcomes are distributed according to the distribution $\mathbf{\tilde{v}}(\strategy_\resident,\strategy_\mutant)$, as described in Proposition~\ref{proposition:last_round}.  
\item The learner's last interaction was not with the role model, with probability $(N\!-\!2)/(N\!-\!1)$. In this case, there are four different subcases, depending on whether the resident's last interaction partner was a mutant or a resident, and depending on whether the mutant's last interaction partner was a mutant or a resident. As a result, the resident's last one-shot payoff is distributed according to $\mathbf{\tilde{v}}(\strategy_\resident,\strategy_\mutant)$ or $\mathbf{\tilde{v}}(\strategy_\resident,\strategy_\resident)$; the mutant's last payoff is distributed according to $\mathbf{\tilde{v}}(\strategy_\mutant,\strategy_\mutant)$ or $\mathbf{\tilde{v}}(\strategy_\mutant,\strategy_\resident)$, respectively. 
\end{enumerate}

\noindent
Let $x(u_\resident,u_\mutant)$ denote the probability that the resident and the mutant received the payoff $u_\resident$ and $u_\mutant$ in their respective last interaction.  
By taking into account the above two cases, we can compute this probability as
\begin{equation}\label{eq:Chi} 
\setlength{\arraycolsep}{1pt} 
\begin{array}{llll}
x(u_\resident,u_\mutant)	 &=
&\frac{1}{N\!-\!1}\cdot  &\tilde v_{u_\resident}\!(\strategy_\resident,\strategy_\mutant)\cdot 1_{(u_\resident,u_\mutant)\in \mathcal{U}_F}\\[0.5cm]
&+	
&\frac{N\!-\!2}{N\!-\!1} \cdot 
&\left[ \frac{k\!-\!1}{N\!-\!2}\frac{k\!-\!2}{N\!-\!3}\cdot \tilde v_{u_\resident}\!(\strategy_\resident,\strategy_\mutant) \,\tilde v_{u_\mutant}\!(\strategy_\mutant,\strategy_\mutant) + 
 \frac{k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!1}{N\!-\!3}\cdot \tilde v_{u_\resident}\!(\strategy_\resident,\strategy_\mutant) \,\tilde v_{u_\mutant}\!(\strategy_\mutant,\strategy_\resident)\right.\\[0.5cm]
&&&\left. +\frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{k\!-\!1}{N\!-\!3} \cdot \tilde v_{u_\resident}\!(\strategy_\resident,\strategy_\resident) \, \tilde v_{u_\mutant}\!(\strategy_\mutant,\strategy_\mutant) + 
 \frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!2}{N\!-\!3} \cdot \tilde v_{u_\resident}\!(\strategy_\resident,\strategy_\resident) \, \tilde v_{u_\mutant}\!(\strategy_\mutant,\strategy_\resident)\right].
\end{array}
\end{equation}

@@

{\bf Motivation.} 
In the limited memory setting, individuals take into account the outcome of one round, and of one interaction. 
This setting can be generalized such that an individual
considers \(m\) rounds and of \(n\) interactions. Here we discuss the case that
the update depends on the last round of \(n\) interactions.

@@

We assume that player the resident receives the payoff \(u_{\resident1}\) with their first interaction and \(u_{\resident 2}\) with their second.
Similarly, the mutant receives payoffs \(u_{\mutant 1}\) and \(u_{\mutant 2}\). 
Let \(x(u_{\resident1} u_{\resident1}, u_{\mutant 1} u_{\mutant 2})\) denote the probability that the resident and the mutant received payoffs  \(u_{\resident1}\), \(u_{\resident2}\) and \(u_{\mutant 1}\), \(u_{\mutant 2}\) in the last round of their respective last two interactions.
This probability can be computed as follows, 
\begin{equation} \label{Eq:xTwoInteractions} 
  \setlength{\arraycolsep}{0.5cm}
  \footnotesize
  \hspace*{-0.8cm} 
  \begin{array}{lll}
  \multicolumn{3}{l}{x(u_{\resident1}u_{\resident 2}, u_{\mutant 1}u_{\mutant 2})= \displaystyle \frac{1}{N\!-\!1}\cdot \tilde{v}_{u_{\resident1}}(\strategy_\resident,\strategy_\mutant)\cdot 1_{(u_{\resident1},u_{\mutant 1})\in \mathcal{U}_F} \cdot A }\\[0.4cm]

  &\multicolumn{2}{l}{+ \frac{N-1}{N-2} \biggl[
  \frac{(k - 1)(k - 2)}{(N\!-\!2)(N\!-\!3)}\,
  \tilde{v}_{u_{\resident1}}\!(\strategy_\resident,\strategy_\mutant) \, \tilde{v}_{u_{\mutant 1}}\!(\strategy_\mutant,\strategy_\mutant) 
  \left( \frac{\tilde{v}_{u_{\resident 2}}\!(\strategy_\resident,\strategy_\mutant)}{N-2} \, 1_{(u_{\resident 2},u_{\mutant 2})\in \mathcal{U}_F} +\frac{N-3}{N-2} [B_1 \!+\! B_2 \!+\! B_3 \!+\! B_4]\right)}\\[0.4cm]

>>>>

\noindent
This document provides further details on our methods and derivations, and it contains additional simulation results. 
Section~\ref{section:model} summarizes the model. 
In particular, we provide further details on our implementation of the evolutionary dynamics, and our use of the rare-mutation limit. 
In Section~\ref{section:analyticalresults}, we derive analytical results for the various settings we consider.
These settings differ in what kind of payoff information individuals take into account when updating their strategies. 
In the perfect-memory setting, individuals take into account all their interactions against all co-players.
In the limited-memory setting, they only consider the very last round of their very last interaction.  
In addition, we describe several model extensions in which the amount of information taken into account is in between these two extremes. 
Finally, Section~\ref{section:furthersimulations} presents further simulation results. 
In particular, we confirm that our main results continue to hold ({\it i}) when mutations are no longer rare, and  ({\it ii}) when players use memory-one strategies instead of reactive strategies.

@@

\noindent
{\bf Summary of the model.} As described in the main text, we study cooperative behavior in a population of size $N$, with $N$ being even.
The dynamics unfold on two time scales. 
The short time scale describes the game dynamics. 
Here the $N$ individuals are randomly matched to form $N/2$ pairs to interact in a repeated prisoner's dilemma.  
Each round, individuals can choose whether to cooperate ($C$) or defect ($D$). 
In the most general setting, the resulting one-shot payoffs can be summarized by the payoff matrix 
\begin{align}
	\bordermatrix{%
		& C & D \cr
		C &\ R &\ S\  \cr
		D &\ T &\ P\ \cr
	} . \label{eq:prisoner}
\end{align}
Here, $R$ is interpreted as the reward for mutual cooperation, $S$ is the sucker's payoff, $T$ is the temptation, and $P$ is the punishment payoff~\citep{axelrod1981evolution}. 
Throughout this work, we parametrize these payoffs as $R\!=\!b\!-\!c$, $S\!=\!-c$, $T\!=\!b$, and $P\!=\!0$, where $b$ and $c$ are the benefit and cost of cooperation, respectively, with $b\!>\!c\!>0$. 
After each round, players learn their co-player's previous action. 
Then the game continues for another round with probability~$\delta$.
Players make their decisions whether to cooperate in any given round based on their reactive strategies $\mathbf{s}\!=\!(y,p,q)$. 
The entry $y$ determines a player's first-round cooperation probability. 
The other entries $p$ and $q$ determine the player's cooperation probability in all subsequent rounds. 
The probability is $p$ if the co-player cooperated in the previous round, and it is $q$ if the co-player defected. 
On this short time scale, the players' strategies are fixed, and players are consecutively matched to play many repeated games with randomly changing interaction partners.

@@

The long time scale describes the evolutionary dynamics. 
Here, players are allowed to update their strategies based on the payoffs they yield. 
We model these strategy updates with a pairwise comparison process~\citep{traulsen2007pairwise}.
This process assumes that at regular time intervals, one player is randomly selected from the population.
We refer to this player as the `learner' (\learner). 
The learner is then given an opportunity to update its strategy. 
There are two possibilities for how this update may occur. 
With probability $\mu$, the player's strategy mutates randomly. 
In that case, the player's new strategy is drawn uniformly from the space of all reactive strategies~$[0,1]^3$.  
With probability $1\!-\!\mu$, the player compares itself with a neighbor. 
In that case, the player randomly picks another individual from the population (referred to as the `role model', \rolemodel). 
The learner adopts the role model's strategy with a probability \(\rho\), given by
\begin{equation} \label{Eq:rho}
    \rho(\pi_{\learner}, \pi_{\rolemodel}) = \frac{1}{1\!+\! e^{\!-\!\beta (\pi_\rolemodel - \pi_\learner)}}.
\end{equation}
The parameter \(\beta\) is the selection strength.
It determines how important payoff differences are for the learner's decision to imitate the role model. 
The variables $\pi_\rolemodel$ and $\pi_\learner$ refer to the relevant payoffs of the role model and the learner, respectively. 
The exact value of these payoffs depend on the players' memory. 
We say players have {\it perfect memory} when $\pi_\rolemodel$ and $\pi_\learner$ are given by the players' expected payoffs (across all rounds and across all possible co-players). 
We say players have {\it limited memory} when $\pi_\rolemodel$ and $\pi_\learner$ are given by the players' realized payoff in the very last round of the game with their very last interaction partner. 
In addition, we consider several model extensions in which individuals have memory capacities in between these two extremes. 
We provide a detailed description of these different settings and the resulting payoffs in Section~\ref{section:analyticalresults}.\\

%% MODEL: Intro to the rare-mutation limit %%

\noindent
{\bf Evolutionary simulations for the rare-mutation limit.} 
To simulate the evolutionary dynamics of the pairwise comparison process, it is sometimes useful to assume that mutations are rare, $\mu\!\rightarrow\!0$. 
In that case, whenever a mutant strategy appears, it either fixes in the population or goes extinct before the next mutant appears. 
As a result, at any given time there are at most two different strategies present in the population~\citep{fudenberg:JET:2006,wu:JMB:2012,mcavoy:jet:2015}. 
This assumptions makes computations more efficient, and it makes some of the results easier to interpret.
In the following, we describe our implementation of the process in the rare-mutation limit in more detail.

%% MODEL: Implementation of the rare-mutation limit %%

Initially, the process starts with a population where all members use the same strategy (referred to as the resident strategy, \resident). 
Then one individual adopts a mutant strategy selected uniformly at random from the set of feasible strategies.
The fixation probability \(\varphi_{\mutant}\) of the mutant strategy can be calculated explicitly~\citep{nowak:Nature:2004},
\begin{equation}\label{eq:fixation_probability}
    \varphi_{\mutant} = \frac{1}{1+\sum\limits_{i=1}^{N-1}\prod\limits_{k=1}^i \frac{\lambda^-_k}{\lambda^+_k}}.
\end{equation}
Here, the index $k$ corresponds to the current number of players with the mutant strategy (mutants). 
The variables \(\lambda^-_k, \lambda^+_k\) are the probabilities that the number of mutants decreases or increases within a single updating step. 
These probabilities depend on the probability that a mutant and a resident are chosen as the learner and the resident, respectively. 
In addition, they depend on the respective switching probability~$\rho$, as described by Eq.~\eqref{Eq:rho}. 
We specify the exact values of  \(\lambda^-_k, \lambda^+_k\) for each memory-setting in the next section.

Depending on the fixation probability \(\varphi_{\mutant}\), the mutant strategy either fixes (becomes the new resident) or goes extinct. 
Afterwards, another random mutant strategy is introduced into the population. 
We iterate this elementary population updating process for a large number of mutant strategies. 
At each step, we record the current resident strategy and the resulting average cooperation rate.

We consider this limit of rare-mutation throughout the main text. 
The respective process is summarised by Algorithm~\ref{algorithm:pairwise_comparison}.
In Section~\ref{section:furthersimulations}, we present additional simulation results to show that our qualitative results continue to hold when the mutation rate is strictly bounded away from zero.

@@

Using the auxiliary notation $r_i:=p_i\!-\!q_i$, we obtain

@@

~\\
{\bf Stochastic stability of cooperation.}
As an application of this formalism, we can compute when cooperation is stochastically stable in the perfect-information setting. 
To this end, suppose there is only a single mutant, $k\!=\!1$. The residents adopt Generous Tit-for-Tat, $\gtft = (1, 1, q)$ and the mutant adopts $\alld\! = \! (0,0,0)$. When two GTFT players interact, the resulting average distribution according to Eq.~\eqref{Eq:vAverage} simplifies to
\begin{align*}
\mathbf{v}(\gtft,\gtft) = (1, 0, 0, 0).
\end{align*}
On the other hand, if \alld{} interacts with \gtft, the respective probabilities become,
\begin{align*}
  \mathbf{v}(\alld,\gtft) = \big(0,~0,~1\!-\!\delta\!+\!\delta q,~\delta(1\! -\! q)\big).
\end{align*}
Based on Eq.~\eqref{Eq:ExpPay}, we can compute the strategies' expected payoffs as
\begin{align*}
\pi_\gtft =  \frac{N\!-\!2}{N-1} (b - c)  \!-\! \frac{1}{N-1}\big(1\!-\!\delta\!+\!\delta q\big)c
\quad \text{and} \quad
\pi_\alld  = \big(1\!-\!\delta\!+\!\delta q\big)b.
\end{align*}
As a consequence, we can calculate the corresponding ratio of transition probabilities according to Eq.~\eqref{eq:LambdaRatio},
\begin{align*}
\frac{\lambda_1^{-}}{\lambda_1^{+}} = e^{-\beta \left( (1-\delta+\delta q)(b+\frac{c}{N-1})- \frac{N-2}{N-1}(b-c) \right)}
\end{align*}
By definition, cooperation is stochastically stable if this ratio exceeds one, which is equivalent to
\begin{equation*}
q<1-\frac{1}{\delta}\cdot \frac{b+(N\!-\!1)c}{(N\!-\!1)b+c}.
\end{equation*}
For such a strategy to be feasible we require $q\!>\!0$, which implies $\delta \!>\! \big(b+(N\!-\!1)c\big)/\big((N\!-\!1)b\!+\!c\big)$.
In particular, in the limit of large populations \(N \!\rightarrow\! \infty\), we obtain that cooperation is stochastically stable if  \(q \!<\! 1\! -\! c/(\delta b)\). 
The minimum continuation probability for such a strategy to exist is $\delta \!>\! c/b$. 
In this way, we recover the classical conditions for cooperation to be feasible under direct reciprocity~\citep{molander:jcr:1985,Nowak1992tit,Schmid:NHB:2021}.

@@

{\bf Computing the distribution of last-round payoffs.}
The case of perfect payoff memory is straightforward to handle; here, every player gets the expected payoff with certainty. 
In comparison, computing transition probabilities for the case of limited payoff memory is more elaborate. 
Here, we need to consider the different possible outcomes that both the learner and role model may have experienced in their very last interaction. 
A further complication arises when the learner's last interaction partner happens to be the role model. 
In that case, the learner's and the role model's last payoff will be correlated (e.g., if the learner got the sucker's payoff of $S$, the role model's payoff is $T$ with certainty). 
To treat the case of limited memory analytically, let $\mathcal{U} \!=\! \{R,S,T,P\}$ be the set of possible one-shot payoffs. 
Since the game ends after round $T$ with probability $\delta^{T}\left(1-\delta\right)$, if two players use reactive strategies $\strategy_1\!=\!(y_1, p_1, q_1$)  and $\strategy_2\!=\!(y_2,p_2,q_2)$, then, by definition, the probability that the first player receives a payoff of $u\!\in\!\mathcal{U}$ in the final round of the game is given by Eq.~\eqref{Eq:vAverage}. To make the strategies explicit, we denote this probability by $v_{u}\!\left(\strategy_{1},\strategy_{2}\right)$.

@@

~\\
\noindent
{\bf Computing the ratio $\lambda^-_k/\lambda^+_k$.}
After these preparations, let us again consider the corresponding population setup, with $N\!-\!k$ residents with strategy  \(\strategy_\resident = (y_\resident, p_\resident, q_\resident)\) and $k$ mutants with strategy  \(\strategy_\mutant = (y_\mutant, p_\resident, q_\mutant)\).
At each step of the evolutionary process we choose a learner and a role model. 
The learner compares the performance of its strategy by comparing its own last one-shot payoff with the last one-shot payoff of the role model. 
In the following, we assume either the learner or the role model is a resident and that the other player is a mutant (otherwise it is certain that the number of mutants does not change). 
There are two major cases to consider.

@@

\begin{enumerate}
\item The learner and the role model have their last respective interaction with each other. This happens with probability $1/(N\!-\!1)$. In that case, there are four possible cases for their joint final payoffs, $(u_\resident,u_\mutant)\!\in\!\mathcal{U}_F\!:=\!\big\{(R,R),\,(S,T),\,(T,S)\,(P,P)\big\}$. These outcomes are distributed according to the distribution $\mathbf{v}(\strategy_\resident,\strategy_\mutant)$ (Eq.~\eqref{Eq:vAverage}).  
\item The learner's last interaction was not with the role model, with probability $(N\!-\!2)/(N\!-\!1)$. In this case, there are four different subcases, depending on whether the resident's last interaction partner was a mutant or a resident, and depending on whether the mutant's last interaction partner was a mutant or a resident. As a result, the resident's last one-shot payoff is distributed according to $\mathbf{v}(\strategy_\resident,\strategy_\mutant)$ or $\mathbf{v}(\strategy_\resident,\strategy_\resident)$; the mutant's last payoff is distributed according to $\mathbf{v}(\strategy_\mutant,\strategy_\mutant)$ or $\mathbf{v}(\strategy_\mutant,\strategy_\resident)$, respectively. 
\end{enumerate}

\noindent
Let $x(u_\resident,u_\mutant)$ denote the probability that the resident and the mutant received the payoff $u_\resident$ and $u_\mutant$ in their respective last interaction.  
By taking into account the above two cases, we can compute this probability as
\begin{equation}\label{eq:Chi} 
\setlength{\arraycolsep}{1pt} 
\begin{array}{llll}
x(u_\resident,u_\mutant)	 &=
&\frac{1}{N\!-\!1}\cdot  & v_{u_\resident}\!(\strategy_\resident,\strategy_\mutant)\cdot 1_{(u_\resident,u_\mutant)\in \mathcal{U}_F}\\[0.5cm]
&+	
&\frac{N\!-\!2}{N\!-\!1} \cdot 
&\left[ \frac{k\!-\!1}{N\!-\!2}\frac{k\!-\!2}{N\!-\!3}\cdot  v_{u_\resident}\!(\strategy_\resident,\strategy_\mutant) \, v_{u_\mutant}\!(\strategy_\mutant,\strategy_\mutant) + 
 \frac{k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!1}{N\!-\!3}\cdot  v_{u_\resident}\!(\strategy_\resident,\strategy_\mutant) \, v_{u_\mutant}\!(\strategy_\mutant,\strategy_\resident)\right.\\[0.5cm]
&&&\left. +\frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{k\!-\!1}{N\!-\!3} \cdot v_{u_\resident}\!(\strategy_\resident,\strategy_\resident) \,  v_{u_\mutant}\!(\strategy_\mutant,\strategy_\mutant) + 
 \frac{N\!-\!k\!-\!1}{N\!-\!2}\frac{N\!-\!k\!-\!2}{N\!-\!3} \cdot v_{u_\resident}\!(\strategy_\resident,\strategy_\resident) \, v_{u_\mutant}\!(\strategy_\mutant,\strategy_\resident)\right].
\end{array}
\end{equation}

@@

{\bf Motivation.} 
In the limited-memory setting, individuals take into account the outcome of one round, and of one interaction. 
This setting can be generalized such that an individual
considers \(m\) rounds and of \(n\) interactions. Here we discuss the case that
the update depends on the last round of \(n\) interactions.

@@

We assume that player the resident receives the payoff \(u_{\resident1}\) with their first interaction and \(u_{\resident 2}\) with their second.
Similarly, the mutant receives payoffs \(u_{\mutant 1}\) and \(u_{\mutant 2}\). 
Let \(x(u_{\resident1} u_{\resident1}, u_{\mutant 1} u_{\mutant 2})\) denote the probability that the resident and the mutant received payoffs  \(u_{\resident1}\), \(u_{\resident2}\) and \(u_{\mutant 1}\), \(u_{\mutant 2}\) in the last round of their respective last two interactions.
This probability can be computed as follows, 
\begin{equation} \label{Eq:xTwoInteractions} 
  \setlength{\arraycolsep}{0.5cm}
  \footnotesize
  \hspace*{-0.8cm} 
  \begin{array}{lll}
  \multicolumn{3}{l}{x(u_{\resident1}u_{\resident 2}, u_{\mutant 1}u_{\mutant 2})= \displaystyle \frac{1}{N\!-\!1}\cdot v_{u_{\resident1}}(\strategy_\resident,\strategy_\mutant)\cdot 1_{(u_{\resident1},u_{\mutant 1})\in \mathcal{U}_F} \cdot A }\\[0.4cm]

  &\multicolumn{2}{l}{+ \frac{N-1}{N-2} \biggl[
  \frac{(k - 1)(k - 2)}{(N\!-\!2)(N\!-\!3)}\,
  v_{u_{\resident1}}\!(\strategy_\resident,\strategy_\mutant) \, v_{u_{\mutant 1}}\!(\strategy_\mutant,\strategy_\mutant) 
  \left( \frac{v_{u_{\resident 2}}\!(\strategy_\resident,\strategy_\mutant)}{N-2} \, 1_{(u_{\resident 2},u_{\mutant 2})\in \mathcal{U}_F} +\frac{N-3}{N-2} [B_1 \!+\! B_2 \!+\! B_3 \!+\! B_4]\right)}\\[0.4cm]

  &&+ \frac{\left(k - 1\right) \left(N - k - 1\right)}{(N\!-\!2)(N\!-\!3)}\,
  v_{u_{\resident1}}\!(\strategy_\resident,\strategy_\mutant) \, v_{u_{\mutant 1}}\!(\strategy_\mutant,\strategy_\resident)  
  \left(\frac{v_{u_{\resident 2}}\!(\strategy_\resident,\strategy_\mutant)}{N-2} \, 1_{(u_{\resident 2},u_{\mutant 2})\in \mathcal{U}_F} + \frac{N-3}{N-2} [C_1 \!+\! C_2 \!+\! C_3 \!+\! C_4]\right) \\[0.4cm]
